---
title: "Calcuating δD of precipitation from δD of leaf waxes"
output: html_document
author: David Fastovich
---

# Introduction

This script takes in $\delta\mathrm{D}_{wax}$ and transforms it into  $\delta\mathrm{D}_{precip}$ in a Bayesian framework informed by the physiology of plant fractionation of hydrogen isotopes. This was approach was first formalized in Tierney et al., (2017, Science Advances) and has since been used a variety of studies that use $\delta\mathrm{D}_{wax}$. Importantly, this procedure assumes that vegetation is largely unchaged over the duration of the time series which can be assessed by plotting the  $\delta^{13}\mathrm{C}_{wax}$ against time.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bayesian Approach

See Tierney et al.,  (2017, Science Advances) for the mathematical basis, but briefly the equations used here:

**Equations to calculate $\delta D_{p}$**

(1) $\delta D_{p} = \frac{1000 + \delta D_{wax}}{\frac{\epsilon_{p}}{1000} + 1} - 1000$

(2) $\epsilon_{p} = f_C4 \times \epsilon_{C4} + (1 - f_{C4}) \times \epsilon_{C3}$

(3) $Y = (\frac{\delta^{13}C_{wax} - \delta^{13}C_{C3}}{\delta^{13}C_{C4} - \delta^{13}C_{C3}}) \times N$

**Prior distrubtion for model parameters ($\theta$; i.e. $f_{C4}$):**

$p(\theta) = \theta^{\alpha - 1}(1 - \theta)^{\beta - 1}$

**Posterior distribution which we will be sampling:**

$p(\theta | Y) \propto \theta^{Y + \alpha - 1} (1 - \theta)^{N - Y + \beta - 1}$

We will be assuming an uninformed prior, hence $\alpha = 1$ and $\beta = 1$

# Resampling Algorithm

1. Sample $C_3$ and $C_4$ $\delta^{13}\mathrm{C}$ end-member values. Here, taken from Sachse et al., (2012, Annu. Rev. Earth Planet. Sci.).

```{r, eval=FALSE}
# First set N - the number of plants contributing to the observed measurement,
# assumed to be large here
N <- 5000

# Value from Garcin et al., (2014, Geochim. Cosmochim. Acta)
d13c_c4_mean <- -19.8
d13c_c4_sd <- 0.4

# Value from Garcin et al., (2014, Geochim. Cosmochim. Acta)
d13c_c3_mean <- -33.4
d13c_c3_sd <- 0.4

# Sample from a normal distribution
d13c_c4_sample <- rnorm(N, mean = d13c_c4_mean, sd = d13c_c4_sd)
d13c_c3_sample <- rnorm(N, mean = d13c_c3_mean, sd = d13c_c3_sd)
```

2. Calculate $Y$ using Equation 3.

```{r, eval=FALSE}
y = ((D13C_WAX_VALUE_GOES_HERE - d13c_c3_sample)/(d13c_c4_sample - d13c_c3_sample)) * N
```

3. Sample the posterior beta distribution.

```{r, eval=FALSE}
# Sample posterior for fraction of C4 plants
# Using sapply here because we want a beta distribution sample for each of the N
# values we calculated in Step 2.
f_c4 <- sapply(Y, function(y) rbeta(1, shape1 = y + 1 - 1, shape2 = N - y + 1 - 1))
```

4. Monte-Carlo resampling to propagate errors from $\epsilon_{C3}$, $\epsilon_{C4}$, and $\delta D_{wax}$ measurements and calculate $\epsilon_{p}$ from Equation 2.

```{r, eval=FALSE}
# Value from Sachse et al., (2012, Annu. Rev. Earth Planet. Sci.)
epsilon_c4_mean <- -126
epsilon_c4_sd <- 4

# Value from Sachse et al., (2012, Annu. Rev. Earth Planet. Sci.)
epsilon_c3_mean <- -113
epsilon_c3_sd <- 2

# Sample epsilon value end members from Sachse et al., (2012, Annu. Rev. Earth Planet. Sci.)
epsilon_c4_sample <- rnorm(1, mean = epsilon_c4_mean, sd = epsilon_c4_sd)
epsilon_c3_sample <- rnorm(1, mean = epsilon_c3_mean, sd = epsilon_c3_sd)

# Sample dD of sample - assuming 2 per mil standard deviation for analytical
# uncertainty.
dD_sample <- rnorm(N, mean = DELTA_D_WAX_VALUE_GOES_HERE, sd = 2)

# Calculate epsilon_p
epsilon_p <- f_c4 * epsilon_c4_sample + (1 - f_c4) * epsilon_c3_sample
```

5. With all model parameters in place, calculate $\delta D_{p}$ from $\delta D_{wax}$ using Equation 1.

```{r, eval=FALSE}
ensemble <- ((1000 + dD_sample) / ((epsilon_p/1000) + 1)) - 1000
```

As written, this code will provide an ensemble of 5,000 estimates of $\delta D_{p}$ for a single pair of $\delta D_{wax}$ and $\delta^{13}C_{wax}$ measurements. In practice, that's now how we want this code to work. We want to feed in an entire spreadsheet and have the results exported in another spreadsheet. Below is a code chunk that can make this happen with data from Tierney et al., (2017, Science Advances).

# Application

All of the code chunks above where set to `eval=FALSE` meaning that they did not run, so I'm coping a lot of that code here. This means that whoever is reading this can simply copy and paste the code chunk below, save it as a .R file, and it will run!

```{r, eval=FALSE}
# Setting seed to make sure the random numbers stay constant for this script
set.seed(1)

# First set N - the number of plants contributing to the observed measurement,
# assumed to be large here.
N <- 5000

# Define the number of iterations (i.e. resamples). Note, I am doing all
# calculations in vector form here, so I never iterate i to the end of iter.
# Rather, I use `iter` to define the number of draws from each distribution
iter <- 1000

# Value from Garcin et al., (2014, Geochim. Cosmochim. Acta)
d13c_c4_mean <- -19.8
d13c_c4_sd <- 0.4

# Value from Garcin et al., (2014, Geochim. Cosmochim. Acta)
d13c_c3_mean <- -33.4
d13c_c3_sd <- 0.4

# Value from Sachse et al., (2012, Annu. Rev. Earth Planet. Sci.)
epsilon_c4_mean <- -126
epsilon_c4_sd <- 4

# Value from Sachse et al., (2012, Annu. Rev. Earth Planet. Sci.)
epsilon_c3_mean <- -113
epsilon_c3_sd <- 2

# Read in data
gc27 <- read.table("~/Downloads/tierney2017gc27.txt", skip = 111, header = TRUE)

# Prepare output list - it will be structured as a list of lists with entries
# 1 to the length of the gc27$dDwax_iv corresponding to each dD_wax sample.
output <- list(
  delta_D_precip = list(),
  f_c4 = list(),
  epsilon_precip = list()
)

# Iterate through each ice-corrected measurement of delta D_{wax}
for(i in 1:length(gc27$dDwax_iv)){
  # Sample d13C end members from Garcin et al., (2014, Geochim. Cosmochim. Acta)
  d13c_c4_sample <- rnorm(iter, mean = d13c_c4_mean, sd = d13c_c4_sd)
  d13c_c3_sample <- rnorm(iter, mean = d13c_c3_mean, sd = d13c_c3_sd)
  
  # Sample epsilon value end members from Sachse et al., (2012, Annu. Rev. Earth Planet. Sci.)
  epsilon_c4_sample <- rnorm(iter, mean = epsilon_c4_mean, sd = epsilon_c4_sd)
  epsilon_c3_sample <- rnorm(iter, mean = epsilon_c3_mean, sd = epsilon_c3_sd)
  
  # Sample dD of sample
  dD_sample <- rnorm(iter, mean = gc27$dDwax_iv[i], sd = 2)
  
  # Calculate y
  Y = ((rep(gc27$d13Cwax[i], iter) - d13c_c3_sample)/(d13c_c4_sample - d13c_c3_sample)) * N
  
  # Sample posterior for fraction of C4 plants
  f_c4 <- sapply(Y, function(y) rbeta(1, shape1 = y + 1 - 1, shape2 = N - y + 1 - 1))
  
  # Monte carlo resampling for epsilon_p
  epsilon_p <- f_c4 * epsilon_c4_sample + (1 - f_c4) * epsilon_c3_sample
  
  # Monte Carlon resampling for dD_precip
  delta_D_precip_ensemble <- ((1000 + dD_sample) / ((epsilon_p/1000) + 1)) - 1000
  
  # Writing to our output list
  output$delta_D_precip[[i]] <- delta_D_precip_ensemble
  output$f_c4[[i]] <- f_c4
  output$epsilon_precip[[i]] <- epsilon_p
}

# Let's get summary statistics from our output list
results <- data.frame(
  dDP = sapply(X = output$delta_D_precip, FUN = mean),
  dDP_1s_lower = sapply(X = output$delta_D_precip, FUN = function(x) mean(x) - sd(x)),
  dDP_1s_lower = sapply(X = output$delta_D_precip, FUN = function(x) mean(x) + sd(x))
)

# And compare against the results that are within the Tierney et al., (2017, 
# Science Advances) data file.
mean(gc27$dDP - results$dDP)
mean(gc27$dDP - results$dDP)
mean(gc27$dDP - results$dDP)

# The difference is 0.01968102 - so its safe to say we've replicated the results
```